---
title: "Country Plastic Waste"
format: 
  dashboard:
    orientation: rows
    theme: flatly
embed-resources: true
logo: plastic-logo.svg
logo-alt: "Recycling Logo with caption"
server: shiny
nav-buttons:
      - icon: github
        href: https://github.com/tylerstoen
        aria-label: GitHub
      - icon: linkedin
        href: https://linkedin.com/in/tyler-stoen
        aria-label: LinkedIn
---

```{r, Packages and Data}
#| code-fold: true
library(tidyverse)
library(gt)
library(rvest)
library(ggplot2)
library(dplyr)
library(httr2)
library(jsonlite)
library(httr)
library(glue)
library(tidyjson)
library(ggridges)
library(scales)
library(stringdist)
library(tidytext)
library(leaflet)
library(maps)
library(plotly)
library(gganimate)
library(gifski)
```

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

```{r}
#| label: data-from-api
#| message: FALSE
#| warning: FALSE
#| output: FALSE

base_url <- "https://restcountries.com/v4/"

restcountries <- GET(glue("{base_url}all?fields=name,continents,region,subregion,unMember,gini,density,geolocation"))

restcountries

restcountries_tidy <- restcountries %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  as_tibble()

```

```{r}
#| label: merge-plastics

library(stringi)

normalize_name <- function(x) {
  x %>% 
    stri_trans_general("Latin-ASCII") %>% 
    str_to_lower() %>% 
    str_remove_all("[^a-z ]") %>% 
    str_trim() %>% 
    str_squish()
}

plastics_clean <- plastics %>% 
  filter(country != "EMPTY") %>% 
  mutate(
    country = case_when(
      country == "Korea" ~ "South Korea",
      country == "Taiwan_ Republic of China (ROC)" ~ "Taiwan",
      country == "United Kingdom of Great Britain & Northern Ireland" ~ 
        "United Kingdom",
      TRUE ~ country
    )
  ) %>%
  mutate(country_key = normalize_name(country))

restcountries_clean <- restcountries_tidy %>% 
  dplyr::select(name.common, name.official, region, subregion,
         density, unMember, geolocation.latitude, 
         geolocation.longitude) %>% 
  mutate(common_key = recode(
    normalize_name(name.common),
    "ivory coast" = "cote divoire"
    ),
        official_key = normalize_name(name.official))
  

# Different name types, try both

joined <- plastics_clean %>%
  left_join(
    restcountries_clean %>% dplyr::select(-official_key),
    by = c("country_key" = "common_key")
  )

matched <- joined %>% filter(!is.na(region))
unmatched <- joined %>% filter(is.na(region)) %>%
  dplyr::select(names(plastics_clean))

refilled <- unmatched %>%
  left_join(
    restcountries_clean %>% dplyr::select(-common_key),
    by = c("country_key" = "official_key")
  )

joined1 <- bind_rows(matched, refilled)
```

```{r}
#| label: merge-plastics-cont
## Fuzzy matching for parent companies using stringdist

noise_words <- c(
  "\\bthe\\b",
  "\\binc\\.?\\b",
  "\\bco\\.?\\b",
  "\\bltd\\.?\\b",
  "\\bcorp\\.?\\b",
  "\\bcompany\\b",
  "\\bcompanies\\b",
  "\\bgroup\\b",       # "Nestlé Group" -> "nestlé"
  "\\bholdings\\.?\\b",
  "\\bllc\\.?\\b",
  "\\bplc\\.?\\b",
  "\\bgmbh\\b",        # common German suffix
  "\\bsa\\b",          # common in Latin America/Europe
  "\\bsrl\\b"
)

normalize_company <- function(x) {
  x %>%
    str_to_lower() %>%
    str_remove_all(paste(noise_words, collapse = "|")) %>%
    str_remove_all("[^a-z0-9 ]") %>%
    str_squish()
}

resolve_typos <- function(names, threshold = 0.1) {
  dist_matrix <- stringdistmatrix(names, names, method = "jw")
  diag(dist_matrix) <- Inf
  
  canonical <- names
  
  for (i in seq_along(names)) {
    best_dist <- min(dist_matrix[i, ])
    if (best_dist < threshold) {
      best_match <- which.min(dist_matrix[i, ])
      canonical[i] <- min(names[i], names[best_match])
    }
  }
  
  tibble(normalized = names, canonical = str_to_title(canonical))
}

companies_raw <- unique(joined$parent_company)

name_map <- tibble(original = companies_raw) %>%
  mutate(normalized = normalize_company(original)) %>%
  left_join(
    resolve_typos(unique(.$normalized)),  # fuzzy match on normalized names only
    by = "normalized"
  )

canonical_labels <- name_map %>%
  count(canonical, original) %>%          # count how often each original appears
  slice_max(n, by = canonical, n = 1, with_ties = FALSE) %>% # pick the most frequent per canonical
  dplyr::select(canonical, display_name = original)

joined <- joined1 %>%
  left_join(name_map %>% dplyr::select(original, canonical),
            by = c("parent_company" = "original")) %>%
  left_join(canonical_labels, by = "canonical") %>% 
  mutate(parent_company = display_name) %>% 
  dplyr::select(-display_name, -canonical) %>% 
  mutate(country = str_to_title(country_key))
```

```{r}
#| label: API, World Bank  API

base_url <- "https://api.worldbank.org/v2/country/all/indicator/"

pop_code <- "SP.POP.TOTL"
gdp_code <- "NY.GDP.MKTP.CD"

pop_url <- str_c(base_url, pop_code, "?format=json&per_page=20000")
gdp_url <- str_c(base_url, gdp_code, "?format=json&per_page=20000")

pop_raw <- GET(pop_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  dplyr::select(
    country.value,
    year = date,
    population = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

gdp_raw <- GET(gdp_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>% 
  pluck(2) %>% 
  as_tibble() %>% 
  dplyr::select(
    country.value,
    year = date,
    gdp = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

pop_and_gdp_raw <- inner_join(pop_raw, gdp_raw, by = c("country.value", "year")) %>% 
  mutate(country.value = recode(
           country.value,
          "Turkiye" = "Turkey",
          "Hong Kong SAR, China" = "Hong Kong",
          "Viet Nam" = "Vietnam",
          "Korea, Rep." = "South Korea",
          "United States" = "United States of America"
          ),
         country_code = normalize_name(country.value))

joined_all <- joined %>% 
  left_join(pop_and_gdp_raw, by = c("country_key" = "country_code", "year")) %>% 
  mutate(
    gdppercapita = gdp / population
  )
                              
```

# All Countries

::: {.callout-note}
## Research Question {height=15%}

How does country geography and economic status effect their amount of recorded plastic waste? Do more developed countries in terms of GDP and population tend to have more or less recorded waste?
:::


## Row {height=85%}

### Tabset 1 {width=50%}

::: panel-tabset

## Country Map

```{r}
world_coords <- map_data("world") |>
  group_by(region) |>
  summarise(
    lng = mean(long, na.rm = TRUE),
    lat = mean(lat, na.rm = TRUE),
    .groups = "drop"
  )

plastics_country <- joined_all |>
  filter(parent_company == "Grand Total") |>
  group_by(country) |>
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    population = first(population),  # adjust if your pop column is named differently
    .groups = "drop"
  ) |>
  mutate(
    plastic_per_capita = total_plastic / population,
    country = recode(
      country,
      "Cote D_ivoire" = "Ivory Coast",
      "ECUADOR" = "Ecuador",
      "Hong Kong" = "Hong Kong",
      "NIGERIA" = "Nigeria",
      "United Kingdom" = "UK",
      "United States of America" = "USA"
    )
  ) |>
  left_join(world_coords, by = c("country" = "region"))

pal <- colorNumeric("Blues", domain = plastics_country$plastic_per_capita)

leaflet(plastics_country |> filter(!is.na(lat), !is.na(lng))) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~lng,
    lat = ~lat,
    radius = ~sqrt(plastic_per_capita) * 200,
    popup = ~paste0(
      "<strong>", country, "</strong><br>",
      "Plastic per Capita: ", round(plastic_per_capita, 4), "<br>",
      "Total Plastic: ", scales::comma(total_plastic), "<br>",
      "Population: ", scales::comma(population)
    ),
    fillOpacity = 0.7,
    stroke = FALSE
  ) |> 
  addControl(
    html = "<div style='background:white; padding:8px; border-radius:5px;
             border:1px solid #ccc; font-size:16px; font-weight:bold'>
             Plastic Waste per Capita by Country
             </div>",
    position = "topright"
  )
```

## Continent Plastic per Capita Distributions

```{r}
#| fig-alt: "A ridgeline density plot showing the log-transformed distribution of plastic waste per capita in pieces per person by continent, with four ridges for Europe, Americas, Asia, and Africa. On the log scale, distributions are approximately bell-shaped and more comparable across continents. Europe has the widest spread and is centered around -10, while Africa is narrower and centered slightly lower. Asia shows a bimodal distribution with two distinct peaks."

vis_1.5 <- joined_all %>%
  filter(!is.na(region), region != "Oceania") %>% 
  group_by(country, region, population) %>%
  summarize(
    total = sum(grand_total, na.rm = TRUE), .groups = "drop"
  ) %>% 
  mutate(
    region = fct_reorder(region, total, .desc = TRUE),
    waste_per_capita = total/population
  ) %>% 
  ggplot(aes(x=log(waste_per_capita), y=region, fill=region)) +
  geom_density_ridges(alpha = 0.5, height = 1, position = "identity") +
  labs(
    title = "Log-Transformed Plastic Waste per Capita by Continent",
    subtitle = "Distribution of Log(Plastic Waste per Capita) by Continent (in Pieces/Person)",
    x = "Log(Plastic Waste per Capita) (Items per Person)",
    y = ""
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 15))
vis_1.5
```

:::


### Tabset 2 {width=50%}

::: panel-tabset

## GDP

```{r}
#| fig-alt: "A scatterplot comparing log-transformed recorded plastic waste on the y-axis against GDP per capita on the x-axis, with points colored by continent and a linear trend line in black. The trend line shows a slight negative relationship, suggesting that countries with higher GDP per capita tend to have slightly lower recorded plastic waste. Most countries cluster at low GDP per capita values. European countries are spread across the full GDP range with high variation in waste levels."

vis_2 <- joined_all %>% 
  filter(!is.na(region), !is.na(gdppercapita)) %>%
  group_by(country, gdppercapita, region) %>% 
  summarize(
    total = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = gdppercapita, y = log(total), color = region)) +
  geom_point(alpha = 1, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7, color = "black") +
  scale_x_continuous(labels = label_dollar()) +
  scale_color_brewer(
    palette = "Dark2",
    name = "World Region"
  ) +
  labs(
    title = "Higher-Income Countries Tend to Record Less Plastic Waste",
    subtitle = "Country-level totals of log-transformed recorded plastic waste vs GDP per capita",
    x = "GDP per Capita (USD)",
    y = "Log(Total Plastic Waste)"
  ) +
  theme_light(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size= 15),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
vis_2
```

## Plastic by Income Bracket

```{r}
# GDP per capita income brackets
income_breaks <- c(0, 10000, 20000, 40000, Inf)
income_labels <- c("Low (<$10K)", "Lower-Middle ($10K-$20K)", 
                   "Upper-Middle ($20K-$40K)", "High (>$40K)")

country_level <- joined_all %>%
  group_by(country, region, gdppercapita, population) %>%
  summarise(total_plastic = sum(grand_total, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    plastic_per_capita = total_plastic / population,
    income_group = cut(gdppercapita, breaks = income_breaks, labels = income_labels)
  )

# Custom function: summary stats for one income group
summarise_income_group <- function(group_name) {
  country_level %>%
    filter(income_group == group_name) %>%
    summarise(
      income_group        = group_name,
      n_countries         = n(),
      avg_gdp_per_capita  = mean(gdppercapita, na.rm = TRUE),
      avg_plastic_per_capita = mean(plastic_per_capita, na.rm = TRUE),
      median_plastic_per_capita = median(plastic_per_capita, na.rm = TRUE),
      total_plastic       = sum(total_plastic, na.rm = TRUE)
    )
}

income_summary <- income_labels %>%
  purrr::map(summarise_income_group) %>%
  list_rbind()

income_summary %>%
  gt() %>%
  cols_label(
    income_group = md("**Income Group**"),
    n_countries = md("**Countries**"),
    avg_gdp_per_capita = md("**Avg GDP per Capita (USD)**"),
    avg_plastic_per_capita = md("**Avg Plastic per Person**"),
    median_plastic_per_capita = md("**Median Plastic per Person**"),
    total_plastic = md("**Total Plastic**")
  ) %>%
  fmt_currency(columns = avg_gdp_per_capita, currency = "USD", decimals = 0) %>%
  fmt_number(columns = c(avg_plastic_per_capita, median_plastic_per_capita), decimals = 5) %>%
  fmt_number(columns = total_plastic, decimals = 0) %>%
  tab_header(
    title = md("**Table 4. Per Capita Recorded Plastic Waste by Economic Group**")
  ) %>%
  tab_source_note(
    source_note = "Countries grouped by GDP per capita using World Bank income thresholds. Recorded plastic per person calculated as total recorded items divided by population."
  )
```

:::

# Country Deep Dive


## Why this country?
We selected Montenegro because it appeared as a strong outlier in several of the plastic waste per capita analyses, with substantially higher values than all other countries in the dataset. This makes Montenegro an interesting case study for exploring how environmental, economic, and demographic indicators relate to waste patterns over time. By focusing on an unusual observation, we can better understand how Montenegro differs from more typical countries represented in the data.



```{r}

renewable_code <- "EG.FEC.RNEW.ZS"  
pop_growth_code <- "SP.POP.GROW"
gdp_pc_code <- "NY.GDP.PCAP.CD"

renewable_url <- str_c(base_url, renewable_code, "?format=json&per_page=20000")
pop_growth_url <- str_c(base_url, pop_growth_code, "?format=json&per_page=20000")
gdp_pc_url <- str_c(base_url, gdp_pc_code, "?format=json&per_page=20000")

# Renewable energy %
renewable_raw <- GET(renewable_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  mutate(
    country = country.value,
    year = as.numeric(date),
    renewable_pct = value
  ) %>% 
  select(country, year, renewable_pct) %>% 
  filter(
    country == "Montenegro",
    year >= 2010,
    year <= 2020
  ) %>% 
  arrange(year)

#pop growth
pop_growth_raw <- GET(pop_growth_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  mutate(
    country = country.value,
    year = as.numeric(date),
    pop_growth = value
  ) %>% 
  select(country, year, pop_growth) %>% 
  filter(
    country == "Montenegro",
    year >= 2010,
    year <= 2020
  ) %>% 
  arrange(year)

#gdp per capita
gdp_pc_raw <- GET(gdp_pc_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>% 
  pluck(2) %>% 
  as_tibble() %>% 
  mutate(
    country = country.value,
    year = as.numeric(date),
    gdp_pc = value
  ) %>% 
  select(country, year, gdp_pc) %>% 
  filter(
    country == "Montenegro",
    year >= 2010,
    year <= 2020
  ) %>% 
  arrange(year)

montenegro <- pop_growth_raw %>%
  left_join(gdp_pc_raw, by = c("country", "year"))

montenegro_full <- montenegro %>%
  left_join(renewable_raw, by = c("country", "year"))

```

##

## Slider plot and check Box  {width="25%"}


```{r}
#| context: setup

library(tidyverse)
library(shiny)
library(httr)
library(jsonlite)
library(scales)
library(ggplot2)
library(leaflet)
library(maps)
library(plotly)
library(shiny)

base_url <- "https://api.worldbank.org/v2/country/MNE/indicator/"

renewable_code <- "EG.FEC.RNEW.ZS"  
pop_growth_code <- "SP.POP.GROW"
gdp_pc_code <- "NY.GDP.PCAP.CD"

get_wb_indicator <- function(code, new_name) {
  url <- str_c(base_url, code, "?format=json&per_page=20000")
  
  GET(url) %>% 
    content(as = "text") %>% 
    fromJSON(flatten = TRUE) %>%
    pluck(2) %>% 
    as_tibble() %>% 
    mutate(
      country = country.value,
      year = as.numeric(date),
      value = as.numeric(value)
    ) %>% 
    select(country, year, value) %>% 
    filter(year >= 2010, year <= 2020) %>%
    rename(!!new_name := value)
}

renewable_raw <- get_wb_indicator(renewable_code, "renewable_pct")
pop_growth_raw <- get_wb_indicator(pop_growth_code, "pop_growth")
gdp_pc_raw <- get_wb_indicator(gdp_pc_code, "gdp_pc")

montenegro_full <- pop_growth_raw %>%
  left_join(gdp_pc_raw, by = c("country", "year")) %>%
  left_join(renewable_raw, by = c("country", "year"))

```

### controls {height="120px"}
```{r}
bslib::layout_columns(
  
  div(
    sliderInput(
      inputId = "years",
      label = "Select Year Range:",
      min = min(montenegro_full$year, na.rm = TRUE),
      max = max(montenegro_full$year, na.rm = TRUE),
      value = c(2010, 2020),
      step = 1,
      sep = ""
    )
  ),

  div(
    checkboxGroupInput(
      inputId = "variables",
      label = "Select Variables to Display:",
      choices = c(
        "Population Growth %" = "pop_growth",
        "GDP per Capita" = "gdp_pc",
        "Renewable Energy %" = "renewable_pct"
      ),
      selected = c("gdp_pc", "renewable_pct")
    )
  ),

  col_widths = c(6, 6)
)
```

## Row

### Montenegro Development Indicators
```{r}
plotOutput("indicator_plot", height = "700px")
```


```{r}
#| context: server

filtered_data <- reactive({
  montenegro_full %>%
    filter(
      year >= input$years[1],
      year <= input$years[2]
    ) %>%
    select(country, year, all_of(input$variables)) %>%
    pivot_longer(
      cols = all_of(input$variables),
      names_to = "indicator",
      values_to = "value"
    ) %>%
    mutate(
      indicator = recode(
        indicator,
        pop_growth = "Population Growth",
        gdp_pc = "GDP per Capita",
        renewable_pct = "Renewable Energy %"
      )
    )
})

output$indicator_plot <- renderPlot({
  filtered_data() %>%
    ggplot(aes(x = year, y = value, color = indicator)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5) +
    facet_wrap(~ indicator, scales = "free_y") +
    scale_color_brewer(palette = "Dark2", guide = "none") +
    labs(
      title = "Montenegro Development Indicators Over Time",
      subtitle = "Indicators update based on selected years and variables",
      x = "Year",
      y = "Value"
    ) +
    theme_light(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
})

```